\section*{Acknowledgments}

This work builds directly on open-source tooling and on local partnerships in
Ouarzazate.

On the software side we thank the authors and maintainers of the Rust, Python
and GIS ecosystems we rely on: the Rust compiler and \texttt{cargo}, crates such
as \texttt{serde}, \texttt{thiserror}, \texttt{tokio}, \texttt{x25519-dalek} and
\texttt{ring} for core protocol and cryptographic primitives, \texttt{reqwest}
and \texttt{warp} for HTTP services, and \texttt{prost} for binary framing.
We also acknowledge the OpenTimestamps project for the timestamping protocol and
reference implementation used to anchor day Merkle roots.

On the security side, we are indebted to the OWASP community for freely available
guidance on web and API hardening, which informed the gateway threat analysis
and defensive measures.

On the data and geospatial side we acknowledge \texttt{sqlite} and
\texttt{SpatiaLite} for embedded storage, GDAL/OGR and QGIS for map production
and spatial analysis, and the broader open-source communities that maintain
these projects.
On the local side we are indebted to the khettara guardians, farmers and
craftspeople in Ouarzazate who allowed test installations and answered detailed
questions about water management, as well as the municipal and heritage staff
who helped align deployments with conservation constraints.
Their feedback on
practical risks, access patterns and acceptable interventions shaped the pod
design, field protocols and choice of measurement sites.

Any errors or omissions in this report remain the responsibility of the author.
