\section{System Architecture}\label{sec:system-architecture}

\subsection{Overview}\label{subsec:overview}

TrackOne implements a three-tier architecture: \textbf{Pod} (sensor + radio) $\rightarrow$ \textbf{Gateway} (verify + batch + anchor) $\rightarrow$ \textbf{Verification} (independent auditor).
The system is designed for offline-tolerant operation with eventual Bitcoin timestamping.

\textbf{Core dataflow:}
\begin{enumerate}[nosep]
\item Pods generate telemetry and transmit authenticated frames over a constrained link (\ac{LoRa} / serial).
\item The gateway validates frames (authenticity + anti-replay) and writes \emph{canonical facts}.
\item Canonical facts are deterministically batched into a per-day artifact (\texttt{day.bin}) and a Merkle root commitment.
\item The day artifact (or its hash) is anchored via OpenTimestamps; proofs are upgraded until they include a Bitcoin attestation.
\item Independent verifiers recompute evidence from canonical facts and validate the anchoring proof.
\end{enumerate}

\paragraph{Ledger vs. projection (clean-room separation)}
The authoritative record is the append-only \emph{ledger} (canonical facts + day artifacts).
Any database, dashboard, or standards \ac{API} (e.g., \ac{OGC} SensorThings) is treated as a \emph{projection} of that ledger.
This separation is deliberate: the ledger contains only accepted, replay-checked facts; rejected or malformed frames are visible only in audit logs and never enter the Merkle set for a day.

% ---------------------------------------------------------------------------
% Graphviz figures (rendered from DOT to PDF/PNG).
% Sources live in src/figs/*.dot.

% ---------------------------------------------------------------------------
% Figure C2 – Pod electronics block diagram
% TODO: Draw simple boxes: PV/supercap/harvester \to 3V3 rail; STM32L0 MCU; SX1276 LoRa; Bio-sensing front-end; RH/T sensor \to Gateway.
% Drop the image at: figs/pod_block_diagram.png
% Placement: Architecture / System Design.

\subsection{Component Responsibilities}
\label{subsec:arch-responsibilities}

This section describes responsibilities at the \emph{architectural contract} level (what each component must ensure), rather than listing implementation scripts.

\paragraph{Pod (device)}
\begin{itemize}[nosep]
\item Samples sensors and forms \emph{environment facts} (\texttt{EnvFact}) with \emph{phenomenon time} semantics (start/end for summaries).
\item Emits authenticated frames containing a monotonic \texttt{frame\_counter}.
\item Operates under tight energy constraints via duty-cycling (transmit, then optional short RX window).
\end{itemize}

\paragraph{Gateway (ingest + ledger)}
\begin{itemize}[nosep]
\item Verifies authenticity (\ac{AEAD}) and enforces anti-replay.
\item Writes canonical facts (append-only) and maintains minimal mutable state required for replay enforcement.
\item Batches accepted facts into deterministic per-day artifacts and produces Merkle commitments.
\item Performs anchoring (OpenTimestamps stamp + upgrade) and retains proofs for independent verification.
\end{itemize}

\paragraph{Verifier (independent audit)}
\begin{itemize}[nosep]
\item Rebuilds Merkle roots from canonical facts and checks them against day commitments.
\item Verifies cryptographic timestamp evidence (\ac{OTS} proofs and/or \ac{TSA} proofs if enabled).
\item Does not require trust in gateway runtime, only in the published artifacts and cryptographic assumptions.
\end{itemize}

\subsection{Time Semantics}
\label{subsec:arch-time}

TrackOne distinguishes two times:
\begin{itemize}[nosep]
\item \textbf{phenomenonTime}: when the observed phenomenon occurred (in \texttt{EnvFact.phenomenon\_time\_start/end}).
\item \textbf{resultTime / ingest\_time}: when the gateway accepted the fact into the ledger (in \texttt{Fact.ingest\_time}).
\end{itemize}
This separation allows pods to operate with weak or absent RTCs while the gateway defines day boundaries for batching.

\subsection{Rust Workspace and Pipeline Boundaries}
\label{subsec:arch-rust-boundaries}

TrackOne is organised as a multi-crate Rust workspace, with clear layering between protocol primitives, embedded firmware, and gateway integration:
\begin{itemize}[nosep]
\item \texttt{trackone-core}: no\_std-capable core (fact schema, \ac{AEAD} frame helpers, Merkle helpers, deterministic commitment serialization).
\item \texttt{trackone-pod-fw}: bare-metal firmware crate (STM32-class targets), reusing \texttt{trackone-core} types and framing logic.
\item \texttt{trackone-gateway}: host-side integration (via PyO3/maturin where appropriate) that uses \texttt{trackone-core} for hot loops.
\item \texttt{trackone-constants}: shared constants (e.g., \texttt{MAX\_FACT\_LEN}) to keep sizing policies consistent.
\end{itemize}
This split lets verification logic remain stable while implementation languages and deployment envelopes evolve.

\subsection{Anchoring Variants}\label{subsec:anchoring-variants}
Beyond OpenTimestamps, deployments MAY enable parallel \ac{RFC}~3161 \ac{TSA} anchoring:
\begin{itemize}[nosep]
\item Daily Merkle root submitted to one or more \ac{TSA} endpoints (\ac{RFC} 3161)
\item Store returned \texttt{.tsr} responses alongside \texttt{.ots} proofs
\item Verification tools can accept either \ac{OTS} or \ac{TSA} proofs for timestamp attestation; peer signatures provide immediate provenance even before blockchain inclusion
\end{itemize}

\subsubsection{Stationary Calendar as an Architectural Component}
The stationary calendar is treated as an architectural component (not merely a test trick): it provides a local, deterministic calendar endpoint for development and controlled integration tests.
It does not replace on-chain verification; it improves pipeline resilience by decoupling day artifact production from public calendar availability.

% ---------------------------------------------------------------------------
% NOTE: Legacy script-level pipeline descriptions (pod_sim.py, frame_verifier.py,
% merkle_batcher.py, ots_anchor.py, verify_cli.py) were intentionally removed
% from this chapter to keep it architectural. Those details belong in developer
% documentation and code comments, not in the system architecture narrative.

\iffalse
% Reviewer-slim version: omit embedded validation details.
\subsection{Embedded Validation Architecture}
The embedded validation stack ensures that the on-pod implementation is wire-compatible with gateway tooling:
\begin{itemize}[nosep]
\item Host and QEMU stress harnesses validate bounded stack usage and a fixed maximum frame size budget.
  \item Golden corpora (facts + frames) are used to assert that independent implementations produce identical commitments.
\end{itemize}
\fi\label{subsec:embedded-validation-architecture5}\label{subsec:embedded-validation-architecture}\label{subsec:embedded-validation-architecture4}\label{subsec:embedded-validation-architecture3}\label{subsec:embedded-validation-architecture2}\label{subsubsec:embedded-validation-architecture}

\subsection{Security Assumptions}\label{subsec:security-assumptions}
\begin{itemize}[nosep]
\item \textbf{Pod security:} Pods may be physically accessible; keys should be provisioned to limit blast radius.
\item \textbf{Gateway trust:} Gateway may buffer and store facts, but cannot forge cryptographic timestamp evidence.
\item \textbf{Anchoring security:} Bitcoin attestation provides a high-cost global timestamp reference.
\item \textbf{Calendars:} Multiple independent calendar endpoints reduce single-point failure.
\end{itemize}

% Reviewer-slim version: omit SensorThings and Kubernetes operational detail.
\subsection{Interoperability via OGC SensorThings}\label{subsec:interoperability-via-ogc-sensorthings}
Beyond internal TrackOne facts, the gateway projects verified observations into a standards-compliant \ac{API}.
We adopt the \ac{OGC} \emph{SensorThings} data model as the public-facing abstraction for telemetry and site metadata~\cite{ogc_sensorthings_part1,ogc_sensorthings_part2}.

\begin{itemize}[nosep]
\item Each physical pod (\texttt{pod\_id}) is modelled as a SensorThings \emph{Thing} with a \emph{Location} in the site \ac{GIS} context.
\item Each measured quantity becomes a separate \emph{Datastream}, harmonized with \texttt{SampleType} and (out-of-band) \texttt{SensorCapability} metadata.
\item Each TrackOne \texttt{Fact} of kind \texttt{Env} maps to a SensorThings \emph{Observation}: \texttt{EnvFact.phenomenon\_time\_start/end} define \emph{phenomenonTime}, while \texttt{Fact.ingest\_time} maps to \emph{resultTime}.
\end{itemize}

SensorThings remains a projection: it can be rebuilt from the ledger and does not participate in Merkle commitments.

\subsection{Local Kubernetes Topology}
\label{subsec:k8s-topology}

A local Kubernetes namespace is used as an \emph{executable topology sketch} for development and integration testing.
The goal is not to prescribe a production orchestration platform, but to keep component boundaries and dependencies explicit and testable.

The local overlay (\texttt{deploy/k8s/local/overlays/local}) expresses:
\begin{itemize}[nosep]
\item \textbf{Long-running services:} gateway, a stationary \ac{OTS} calendar, and a database used for projections.
\item \textbf{Persistent volumes:} for facts/proofs (append-only artifacts) and for database state (rebuildable view).
\end{itemize}

By making the topology executable, the deployment becomes a living research harness rather than a one-off diagram.
It is designed to scale into future work: multi-gateway pilots, redundant anchors, monitoring and alerting, automated verification jobs, and integration with organization-level controls can be added as new services or overlays without changing the core ledger semantics.

\subsection{HIL Mediation via Kubernetes}\label{subsec:hil-mediation-via-kubernetes}
\textbf{Intent:} keep real-time hardware-in-the-loop (HIL) control deterministic and outside the cluster, while using Kubernetes to orchestrate experiments, capture artifacts, and make results reproducible and auditable.
This aligns the lab pipeline with the same verifiable evidence flow used in field pilots.

\textbf{Procedure (reference flow):}
\begin{enumerate}[nosep]
  \item \textbf{Configure the HIL rig:} LabVIEW (or an equivalent RT harness) exposes a control \ac{API} and a data export path; hardware timing stays local.
  \item \textbf{Launch from k8s:} a Job triggers a run with a versioned config (git \ac{SHA}, hardware profile, cadence policy).
  \item \textbf{Capture outputs:} raw measurements and logs are collected into append-only artifacts (facts, proofs, and metadata).
  \item \textbf{Verify and anchor:} TrackOne verification jobs recompute commitments and produce day roots; anchors are created as in the field flow.
  \item \textbf{Publish results:} projections update the database for dashboards and reports, while the canonical artifacts remain the source of truth.
\end{enumerate}

\paragraph{Vertical integration scope.}
TrackOne is designed as a vertically integrated system spanning from
hardware design (PCB, energy harvesting, radio selection) through
firmware (bare-metal Rust on STM32-class \ac{MCU}s), cryptographic protocol
(\ac{AEAD} framing, key agreement, anti-replay), application-layer commitment
(canonical facts, Merkle batching, \ac{OTS} anchoring), and organizational
governance alignment (\ac{NIST} C\ac{SF}~2.0 mapping).
This vertical span is
deliberate: integrity assurance for long-lived telemetry requires that
no layer silently degrades trust.

This scope statement is \emph{not} an \ac{OSI}-layer mapping.
TrackOne deliberately controls only the tiers relevant to evidence integrity;
it does not claim coverage of all seven \ac{OSI} layers.
Table~\ref{tab:osi-scope} clarifies exactly where TrackOne's responsibilities
start and stop.

\begin{table}[H]
\centering
\caption{TrackOne scope mapped against OSI layers (for reviewer orientation only).}
\label{tab:osi-scope}
\small
\begin{tabular}{@{}clll@{}}
\toprule
\textbf{OSI Layer} & \textbf{TrackOne's actual touch} & \textbf{Ownership} \\
\midrule
7 --- Application   & Canonical facts, Merkle batching, \ac{OTS} anchoring, SensorThings projection & \checkmark\ Strong \\
6 --- Presentation  & \ac{CBOR}/\ac{JSON} canonicalization, \ac{AEAD} envelope                                & \checkmark\ Strong \\
5 --- Session       & One-shot provisioning key exchange, frame-counter window                  & $\circ$\ Partial \\
4 --- Transport     & Test harness uses UDP; production \ac{LoRa} bypasses L4                        & $\times$\ Out-of-scope \\
3 --- Network       & No IP routing; pods are address-free                                      & $\times$\ Out-of-scope \\
2 --- Data Link     & Configure SX1276 \ac{LoRa} MAC but do not replace it                          & $\circ$\ Partial \\
1 --- Physical      & PCB, antenna, power budget, radio selection                               & $\circ$\ Partial \\
\bottomrule
\end{tabular}
\end{table}

The absence of layers~3--4 is not a gap---it is a design choice that removes
the attack surface and keeps the trusted computing base minimal in constrained
environments.
Where routable IP \emph{is} required (e.g., gateway back-haul), a
complementary WireGuard tunnel supplies encrypted L3 transport without
altering the evidence pipeline
(see~\S\ref{subsec:layering-wireguard}).
% ---------------------------------------------------------------------------
% NOTE: We intentionally omit YAML listings and kubectl invocations here.
% The architecture chapter should describe contracts and boundaries, not
% operational steps. Those details have a home in deploy/ and docs/.

\subsection{Architecture Decision Matrix}\label{subsec:architecture-decision-matrix}
To help visualise how major components defend different system goals, Figure~\ref{fig:risk-opportunity-matrix} provides a likelihood-by-impact matrix used during review.
