\section{Conclusion and Roadmap}

\subsection{M\#4 Achievements (TrackOne as Track One of Three)}

Track1 (Barnacle Sentinel) has completed Milestone \#4, delivering a production-informed, verifiable telemetry pipeline with Bitcoin-anchored timestamp proofs. In the framing of ADR-022, this is \emph{Track One}: a hardened evidence pipeline rather than a full control system.

\textbf{Evidence and integrity pipeline:}
\begin{itemize}[nosep]
    \item XChaCha20-Poly1305 AEAD with X25519/HKDF-SHA256 and a 64-frame replay window provide strong confidentiality and integrity for framed uplink telemetry, with persistent state across gateway restarts.
    \item Canonical JSON and hash-sorted Merkle trees (ADR-003) make day roots reproducible by independent verifiers regardless of ingestion order.
    \item Daily OpenTimestamps anchoring to Bitcoin, plus a verification CLI, allow third parties to recompute roots from \texttt{facts/} and check proofs without trusting the gateway operator.
\end{itemize}

Together, these results validate the core thesis: resource-constrained subterranean pods can produce telemetry whose \emph{integrity and timing} are publicly auditable, even if later control logic and safety interlocks are implemented in other tracks.

\subsection{Roadmap: M\#5 and Beyond}

We organise follow-up work into near-term hardening of TrackOne and longer-term tracks for control and performance.

\textbf{Near-term hardening (M\#5, TrackOne scope):}
\begin{itemize}[nosep]
    \item \textbf{Ledger and operator UX:} lightweight ``ledger'' views (CLI or web) for auditors to browse blocks, day records, proofs, and peer attestations, driven by a simple JSON API.
    \item \textbf{Automated OTS workflow:} encode the manual anchor/upgrade steps into a small, testable job (cron or systemd) with clear states (pending, anchored, upgraded) and hooks for notifications.
    \item \textbf{Outage and gap logging:} basic outage logger tracking expected pod cadence and surfacing missing data days as first-class events, aligned with the operational constraints of ADR-022.
    \item \textbf{Multi-day verification:} convenience tooling to verify \texttt{prev\_day\_root} chains over multi-day windows and report breaks or ambiguous anchors.
\end{itemize}

\textbf{Longer-term enhancements (Track Two / Three and M\#6+):}
\begin{itemize}[nosep]
    \item \textbf{Key rotation and attestation:} realise ADR-001 and ADR-006 with scheduled key rotation and Ed25519-based device attestation, including revocation workflows.
    \item \textbf{Air-gapped and intermittent deployments:} codify offline verification procedures (signed header bundles, removable media workflows) for sites that cannot run full Bitcoin or continuous calendars.
    \item \textbf{Performance and Rust core:} use ADR-011 benchmarks to decide when to introduce Rust/PyO3 hot paths (ADR-017) for Merkle and AEAD without changing the Python-facing API.
    \item \textbf{Parallel anchoring and PKI:} productionise ADR-015 (RFC 3161 TSA alongside OTS) for deployments that require PKI-backed timestamps in addition to Bitcoin-based proofs.
\end{itemize}

\subsection{Operational Considerations (summary)}

Operationally, TrackOne is ready for disciplined pilot deployments under the constraints in Section~\ref{sec:security} and ADR-022:
\begin{itemize}[nosep]
    \item \textbf{Deployment profiles:} cloud-connected gateways can run full Bitcoin + public calendars; intermittently connected gateways batch OTS operations; air-gapped sites rely on manual export/import and signed header bundles.
    \item \textbf{Sizing envelope:} current implementation comfortably supports small-to-medium pod fleets with typical cadences (e.g., one frame per 5 minutes per pod); Merkle computation remains practical for daily fact counts well below $10^6$.
    \item \textbf{Safety net:} CI, schemas, and the safety-net mechanisms documented in Section~\ref{sec:security} must be kept enabled; TrackOne data should not yet be used as the sole basis for autonomous actuation without an additional safety case.
\end{itemize}

\subsection{Research Contributions}

This work presents Track1 as a \textbf{practical, open reference implementation} that integrates well-known cryptographic components and timestamping workflows for a focused deployment niche: subterranean heritage and irrigation sites requiring long-term data verifiability without vendor lock-in.

\textbf{Key insights:}
\begin{enumerate}[nosep]
    \item \textbf{Ultra-low power adaptation:} XChaCha20-Poly1305 AEAD with LoRa-friendly framing and duty-cycle constraints enables multi-year battery operation in typical scalar-telemetry configurations.
    \item \textbf{Deterministic batching:} Canonical JSON and hash-sorted Merkle trees ensure reproducible roots for independent verification, even under non-deterministic ingestion order and partial failures.
    \item \textbf{Negligible marginal-cost anchoring:} Daily OpenTimestamps batching amortizes per-fact blockchain costs; after initial node/setup overhead, the incremental cost is effectively negligible compared with commercial timestamping services.\footnote{Initial Bitcoin node setup and header synchronization are one-time costs; ongoing node maintenance is small and amortized over many facts.}
    \item \textbf{Open implementation and reproducibility:} A documented reference implementation, ADRs, test artifacts, and a LaTeX report enable third-party auditors to reproduce verification results and reason about design trade-offs.
\end{enumerate}

\textbf{Positioning vs. prior art:}
\begin{itemize}[nosep]
    \item \textbf{CT/Trillian:} Track1 adopts Merkle transparency concepts but for IoT telemetry instead of TLS certificates, using Bitcoin as the public anchor rather than dedicated CT log operators.
    \item \textbf{Guardtime KSI:} Provides similar timestamping capabilities under different trust and licensing assumptions; Track1 fills the niche of an open, Bitcoin-backed reference stack.
    \item \textbf{Helium / blockchain-native LoRa:} Focus on network economics and coverage; Track1 instead targets duty-cycle- and power-constrained pods with daily batching and independent verification.
    \item \textbf{Commercial IoT platforms:} Typically centralise telemetry and logging under vendor control; Track1 emphasises auditor independence and forwards-compatibility via open formats and ADRs.
\end{itemize}

\subsection{Final Remarks and Versioning}

Track1 achieves its immediate goal: a verifiability-first evidence pipeline for resource-constrained subterranean pods, with clearly documented limits and operational constraints. Milestone \#4 delivers a solid foundation with reproducible artifacts, ADRs, and a security/safety net that can support pilot deployments under the guardrails of ADR-022.

Short term, the focus is on hardening TrackOne for pilots: keeping test coverage at or above the agreed threshold, adding a minimal ledger UI, and turning the documented operational procedures into maintained runbooks. Medium term, the release cadence favours steady, auditable updates over rapid feature drops: incremental 0.1.x versions as ledger automation and benchmarking (ADR-011) land, followed by 0.2+ milestones for larger shifts (Rust modules, watermarking experiments) once justified by benchmarks and funding.

The overall strategy remains deliberate: ship and maintain Option~A (this evidence pipeline) as a trustworthy base, keep Options~B/C (watermarking, Rust core) explicitly documented in ADRs, and only advance them when the hardware, benchmarks, and safety analysis justify expanding the system's scope beyond TrackOne.
