\section{Cryptographic Design (ADR-001, ADR-002, ADR-005, ADR-018)}
% ADR reference labels for cross-referencing
\newcommand{\adrlabellist}{% helper to avoid repeated label warnings
}
\label{adr:crypto-design}
\label{adr:nonce-policy}
\label{adr:ots}
\label{adr:benchmarking}
\label{adr:rust-pyo3}
\label{adr:qim-watermarking}

\textbf{Implementation:} PyNaCl (libsodium) for all cryptographic operations.

\textbf{Primitives:} X25519 + HKDF-SHA256 + XChaCha20-Poly1305; Ed25519; SHA-256.

% Reference for deterministic vectors location used in tests
\textit{Deterministic crypto vectors: see} \texttt{toolset/unified/crypto\_test\_vectors.json} \textit{(mirrored by tests; ensure path available to avoid skips).}

\subsection{Provisioning}
Ephemeral ECDH using X25519; salt = $H(N_g \| N_p \| T_{pod} \| B)$.

Derive $CK_{\uparrow}$ (uplink), $CK_{\downarrow}$ (downlink) via HKDF-SHA256 (RFC 5869).

Device table schema v1.0 enforces forward-only policy: requires \texttt{\_meta.version = "1.0"}, 8-byte salt (\texttt{salt8}), and 32-byte uplink key (\texttt{ck\_up}).

\subsection{Telemetry frame}
Header: dev\_id(u16), msg\_type(u8), fc(u32), flags(u8); AEAD payload with XChaCha20-Poly1305.

24-byte nonce construction: $\mathrm{nonce} = \mathrm{salt}_8 \| \mathrm{fc}_8 \| \mathrm{rand}_8$ \textit{(see ADR~\ref{adr:nonce-policy} for randomness and nonce policy)}

AAD (additional authenticated data): $\text{dev\_id} \| \text{msg\_type}$ binds header to ciphertext.

\subsection{Replay policy}
Gateway maintains per-device sliding window on frame counter $(fc)$.
Default window size $w=64$.

A frame is accepted if: $0 < \Delta \mathrm{fc} \le w$ relative to \texttt{highest\_fc\_seen}.

Duplicate/stale frames ($\Delta \le 0$) and jumps beyond $w$ are rejected with logged warnings.

\subsection{Milestone progression}
\begin{itemize}[nosep]
  \item \textbf{M\#0:} Merkle batching, canonicalization, OTS anchoring infrastructure
  \item \textbf{M\#1:} Framed ingest with stub AEAD (plaintext ct, placeholder tag) to validate framing and replay
  \item \textbf{M\#2:} Real XChaCha20-Poly1305 AEAD with test vectors and AAD enforcement
  \item \textbf{M\#3:} PyNaCl migration, device table v1.0, comprehensive testing; CI/CD Python 3.11--3.13
  \item \textbf{M\#4 (current):} OTS anchoring finalized; \textbf{151 tests, 85\% coverage}; expanded security and pipeline edge-case tests
  \item \textbf{Post M\#4 (planned):} Ledger UI, outage logger, automated OTS upgrade
\end{itemize}

\subsection{Cryptographic library consolidation (ADR-005)}
M\#3 migrated from \texttt{cryptography} + \texttt{pynacl} to PyNaCl-only:
\begin{itemize}[nosep]
    \item Single dependency for all primitives (X25519, HKDF, ChaCha/XChaCha, Ed25519)
    \item Unified API with libsodium's proven implementations
    \item Deterministic test vectors regenerated with PyNaCl bindings
    \item Simplified maintenance and security updates
\end{itemize}

\subsection{Contrast with watermarking proposals}
Classical watermarking (e.g., Quantization Index Modulation on analog signals) embeds auxiliary bits inside rich media streams. Track1 pods emit scalar telemetry only; there is no image pipeline and no spare energy for transform-domain embedding. Our threat model targets forged facts and replay, not exfiltration via media. Consequently, rather than hiding markers, we:\footnote{Documented in ADR~\ref{adr:nonce-policy} (nonce policy) and ADR~\ref{adr:ots} (OTS anchoring).}
\begin{enumerate}[nosep]
    \item Bind each fact cryptographically (AEAD + canonical JSON + Merkle leaf)
    \item Anchor the batch root publicly (OTS/TSA)
    \item Provide verifiers the ability to recompute and compare roots
\end{enumerate}
Signal-watermark concepts remain catalogued under Option~B as future research when higher-bandwidth biosignals are part of the payload.

\subsection{Benchmarking and Rust roadmap}
ADR~\ref{adr:benchmarking} (Benchmarking Strategy) defines suites to measure cryptographic throughput, Merkle construction time, and OTS latency. Those metrics will drive ADR~\ref{adr:rust-pyo3} (Rust/PyO3 integration) decisions: only when benchmarks show that Python hot paths jeopardize pod duty-cycle budgets do we port modules to Rust. Until then, PyNaCl-backed Python remains the authoritative implementation.
