\section{Related Work}

\subsection{Merkle Tree Timestamping Systems}

\textbf{Certificate Transparency (RFC 6962, 2013):} Google's CT system uses Merkle trees to create append-only logs of TLS certificates with public auditability. CT demonstrates the viability of Merkle-based timestamping at scale (billions of certificates), but targets web PKI, not IoT telemetry.\ CT logs are operated by trusted third parties, whereas Track1 uses Bitcoin blockchain for trustless verification.

\textbf{Trillian (Google, 2017):} Generic Merkle tree implementation extracted from CT. Provides verifiable data structures as a service but requires infrastructure deployment.\ Track1's daily batching is conceptually similar but optimized for LoRa duty cycle constraints and Bitcoin anchoring costs.

\textbf{Guardtime KSI (2007-present):} Commercial blockchain timestamping service used for IoT and critical infrastructure. KSI uses proprietary blockchain (not Bitcoin) with Merkle aggregation. Requires enterprise licensing and closed-source verification.\ Track1 provides equivalent functionality with open-source implementation and public Bitcoin verification.

\subsection{Blockchain-Based IoT Systems}

\textbf{Helium Network (2019-present):} Decentralized wireless network using Helium blockchain (now Solana).\ Provides proof-of-coverage and data credits for IoT devices.\ Charges per-packet fees (\$0.00001/packet), making it unsuitable for multi-year autonomous deployments.\ Track1's daily OTS anchoring provides zero marginal cost after initial stamp.

\textbf{IOTA Tangle (2016-present):} Directed acyclic graph (DAG) designed for IoT micropayments.\ Requires continuous connectivity and computational overhead for tip selection.\ Track1's gateway-based batching enables offline pod operation with intermittent uplinks.

\textbf{Ethereum-based systems:} Various research projects use Ethereum smart contracts for IoT data integrity.\ Gas fees (\$0.50-\$5 per transaction) make per-message timestamping prohibitively expensive.\ Track1's Merkle batching amortizes blockchain costs across 1000+ daily facts.

\subsection{OpenTimestamps Ecosystem}

\textbf{OpenTimestamps Protocol (2016):} Standard for Bitcoin blockchain timestamping.\ Track1 builds on OTS but adapts it for LoRa telemetry workflows: daily batching instead of real-time stamping, gateway-based aggregation, and offline verification with pre-downloaded headers.

\textbf{Existing OTS implementations:} Focus on document timestamping (PDFs, git commits).\ Track1 extends OTS to IoT telemetry with Merkle aggregation, canonical JSON determinism, and day-chaining for temporal continuity.

% Requires in preamble:
% \usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,shapes.geometric}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
  node distance=1.4cm and 1.5cm,
  block/.style = {draw, rounded corners, align=center, minimum width=2.6cm, minimum height=0.9cm, font=\small},
  line/.style  = {-{Latex[length=2mm]}, thick},
  group/.style = {draw, dashed, inner sep=0.3cm, rounded corners, label={[anchor=north west, font=\footnotesize, inner sep=2pt]north west:#1}}
]

% Top row: Local site processing
\node[block] (sensor) {Sensor data\\{\scriptsize(temp, humidity)}};
\node[block, right=of sensor] (canon) {Canonicalize\\\& normalize};
\node[block, right=of canon]  (hash)  {Hash\\{\scriptsize(SHA-256)}};
\node[block, right=of hash]   (merkle){Merkle\\batching};

% Middle row: Global log service
\node[block, below=2.2cm of sensor] (seq) {Sequence\\Integrate};
\node[block, right=of seq] (ckpt) {Checkpoint\\+ witness};
\node[block, right=of ckpt] (pub) {Publish\\checkpoint};
\node[block, right=of pub] (ots) {OTS\\Bitcoin anchor};

% Note between top and middle
\node[below=0.8cm of hash, align=center, font=\scriptsize, text width=5cm] (note)
  {\textit{Raw data stays local; only hashed commitments leave the site.}};

% Artifacts node
\node[block, below=2.2cm of seq] (artifacts) {Public artifacts\\{\scriptsize(checkpoint, proofs)}};

% Bottom row: Verifier operations
\node[block, below=2.2cm of artifacts] (vfetch) {Fetch payload\\+ proofs};
\node[block, right=of vfetch] (vhash) {Recompute\\hash};
\node[block, right=of vhash] (vincl) {Verify Merkle\\inclusion};
\node[block, right=of vincl] (vck) {Verify\\checkpoint};
\node[block, below=1.4cm of ots] (vots) {Verify OTS\\vs Bitcoin};

% Flows - top row
\draw[line] (sensor) -- (canon);
\draw[line] (canon) -- (hash);
\draw[line] (hash) -- (merkle);

% Flow from merkle down to seq
\draw[line] (merkle) -- (merkle |- seq.north);
\draw[line] (merkle |- seq.north) -- (seq);

% Middle row flows
\draw[line] (seq) -- (ckpt);
\draw[line] (ckpt) -- (pub);
\draw[line] (pub) -- (ots);

% Flow from pub to artifacts
\draw[line] (pub) -- (pub |- artifacts.north);
\draw[line] (pub |- artifacts.north) -- (artifacts);

% Flow from artifacts down
\draw[line] (artifacts) -- (vfetch);

% Bottom row flows
\draw[line] (vfetch) -- (vhash);
\draw[line] (vhash) -- (vincl);
\draw[line] (vincl) -- (vck);

% Flow from vck to vots
\draw[line] (vck) -- (vck |- vots.west);
\draw[line] (vck |- vots.west) -- (vots);

% Flow from ots down to vots
\draw[line] (ots) -- (vots);

% Group boxes to emphasize "globality"
\node[group=Local site]        (gLocal) [fit=(sensor)(canon)(hash)(merkle)(note)] {};
\node[group=Global log service](gLog)   [fit=(seq)(ckpt)(pub)(ots)] {};
\node[group=Public networks]   (gPub)   [fit=(artifacts)] {};
\node[group=Verifier anywhere] (gVer)   [fit=(vfetch)(vhash)(vincl)(vck)(vots)] {};

\end{tikzpicture}
\caption{Track1 data flow demonstrating globality: local sensor data is hashed and committed through Merkle trees to public Bitcoin blockchain via OpenTimestamps, enabling independent verification anywhere without exposing raw telemetry.}
\label{fig:globality}
\end{figure}

\subsection{Low-Power IoT Cryptography}

\textbf{TinyCrypt, ARM mbed TLS:} Lightweight crypto libraries for microcontrollers. Demonstrate feasibility of AES-GCM and ECC on constrained devices.\ Track1 uses PyNaCl (libsodium wrapper) for gateway crypto while pods use minimal nonce construction (salt8 + counter).

\textbf{LoRaWAN Security (LoRa Alliance):} Standard AES-128 encryption with session keys.\ Track1 targets non-LoRaWAN deployments (simple LoRa P2P) with XChaCha20-Poly1305 for 256-bit security and larger nonce space.

\subsection{Track1's Unique Contribution}

While individual components (Merkle trees, AEAD, OTS) are established, Track1's contribution is the \textbf{system-level integration} for a specific deployment scenario:

\begin{itemize}[nosep]
    \item \textbf{vs. Certificate Transparency:} Adapted for IoT telemetry instead of TLS certificates; Bitcoin verification instead of trusted log operators
    \item \textbf{vs. Guardtime KSI:} Open-source implementation; public Bitcoin blockchain instead of proprietary KSI chain
    \item \textbf{vs. Helium Network:} Zero marginal cost (daily batching) instead of per-packet fees; ultra-low power budget
    \item \textbf{vs. Ethereum systems:} 1000× cost reduction via Merkle aggregation (\$0.001/day vs. \$1+/day)
    \item \textbf{vs. LoRaWAN:} Stronger crypto (XChaCha20 vs.\ AES-128); public auditability; no network server dependency
\end{itemize}

Track1 targets heritage sites requiring \textbf{multi-decade autonomy} and \textbf{vendor-independent verification}—a niche not addressed by commercial IoT platforms or blockchain systems optimized for real-time transactions.

\section{Limitations and Future Work}

\subsection{Current Limitations (M\#4)}

\textbf{On-pod cryptography:} Current implementation uses gateway-based AEAD encryption.
Pod payloads are encrypted in transit (LoRa frames) but encryption keys are managed by gateway. Future work (M\#6) will implement on-pod key derivation for end-to-end encryption.

\textbf{Key rotation:} ADR-001 specifies key rotation workflows, but automation is not yet implemented. Manual key updates are required for long-term deployments (M\#5 target: automated weekly ratcheting).

\textbf{Replay window size:} Fixed at 64 frames. For high-latency scenarios (weeks offline), larger windows or epoch-based schemes may be needed.

\textbf{Single gateway and state sync:} Current architecture assumes a single gateway per site. Multi-gateway deployments with state synchronization and consistent replay-window enforcement are non-trivial and deferred.

\textbf{OTS upgrade automation:} Proofs currently require an \texttt{ots upgrade} lifecycle step that may be manual in constrained environments. M\#5 will add automation and error handling.

% --- New: explicit safety-oriented framing ---
\subsection{Safety-relevant limitations}
TrackOne improves integrity and auditability of telemetry but does not, in its current deliverable, provide the guarantees required for autonomous actuation in safety-sensitive deployments. Concretely:
\begin{itemize}[nosep]
  \item The system does not guarantee bounded-latency delivery of facts under all field conditions; decision logic that depends on real-time telemetry must be gated by human-in-the-loop checks (see R-3 in Requirements).
  \item Hardware failure modes (sensor drift, electrode corrosion, ADC saturation) are outside the scope of TrackOne's core pipeline and must be mitigated by hardware test plans and periodic calibration procedures.
  \item No formal SIL-level certification exists for control; TrackOne outputs should be treated as auditable evidence, not a certification of safe-to-act status.
\end{itemize}

% --- New: Documentation & Traceability ---
\subsection{Documentation and traceability}
For an end-of-year project intended to be deployed in challenging environments, documentation is not optional. The following artifacts are ``living'' and must be kept in sync with code and deployments:
\begin{itemize}[nosep]
  \item Requirements (including the irrigation-specific requirements added in this paper) with traced links to tests and ADRs.
  \item Architecture diagrams and deployment profiles (per-site) that document pod/gateway/backend interactions and escalation paths.
  \item Test plans and runbooks covering on-pod diagnostics, failure-injection scenarios (link loss, clock drift, storage exhaustion), and recovery procedures.
  \item Known limitations and mitigation checklists attached to each release (what an operator must not do with this release).
\end{itemize}

These artifacts should be versioned alongside software releases and available to field operators and auditors.

% --- New: Operational procedures (concise) ---
\subsection{Operational procedures}
Deploying TrackOne in irrigation contexts requires clear operational playbooks. Minimal recommended procedures:
\begin{enumerate}[nosep]
  \item Site bootstrap: verify provisioning (device table v1.0), perform a 72-hour pilot with local monitoring enabled, and confirm anchor delivery for two consecutive days before enabling any automatic actuation that affects irrigation.
  \item Firmware update: stage updates to a pilot subset of devices for 7 days; monitor telemetry integrity and rollback on anomalous behavior.
  \item Storage exhaustion: when on-device or gateway storage reaches 80\% capacity, system must stop non-critical writes and alert operators; at 95\% capacity it must trigger emergency archival and operator intervention.
  \item Operator override logging: any manual override must create an auditable record that is included in the next day's Merkle root and proof, so that later audits can reconstruct operator interventions.
\end{enumerate}

\subsection{Ethical and regulatory considerations}

\textbf{Fact confidentiality:} Telemetry payloads remain local (gateway storage).\ Only Merkle roots are public (Bitcoin blockchain via OTS). Individual facts cannot be reconstructed from blockchain data.

\textbf{Right to be forgotten (GDPR):} Local fact deletion is possible (gateway operator decision).\ However, Merkle root in Bitcoin blockchain is immutable.\ Heritage sites should assess GDPR applicability before deployment.

\textbf{Metadata leakage:} Frame timing and pod IDs are visible to LoRa eavesdroppers.\ XChaCha20 encryption hides payload content but not traffic patterns.

\subsection{Environmental Impact}

\textbf{Battery disposal:} CR2032 lithium cells require proper recycling.\ Multi-year autonomy reduces disposal frequency vs.\ annually-replaced batteries.

\textbf{Bitcoin energy consumption:} OTS leverages existing Bitcoin infrastructure (no additional mining).\ Daily batching minimizes per-fact environmental impact (1 anchor amortized across 1000+ facts).

\subsection{Heritage Site Stewardship}

\textbf{Reversibility:} Adhesive pod mounts are non-invasive and removable without surface damage.\ Critical for heritage site preservation ethics.

\textbf{Curator control:} Gateway operator decides what data to collect and publish (Merkle roots).\ Public verification enables accountability but respects site autonomy.

\textbf{Long-term access:} Open-source design ensures data remains accessible even if Track1 project discontinues.\ Bitcoin verification survives organizational changes.

\subsection{Responsible Disclosure}

\textbf{Security vulnerabilities:} All cryptographic designs documented in ADRs. Independent security review encouraged (MIT license).\ No security-by-obscurity.

\textbf{Test coverage:} ~85\% coverage with 181 tests provides confidence but not proof of correctness.\ Formal verification out of scope for M\#4.

\textbf{Known attack vectors:} Physical pod tampering, gateway compromise, replay attacks beyond window size.\ Documented in ADR-002 threat model.

\subsection{Stakeholder expectations and ethical scope management}
Field stakeholders occasionally push comfort-zone technologies (e.g., generic image watermarking or steganography) even when they do not match the telemetry problem space. Heritage pods emit scalar facts; there are no images today. Implementing watermarking to appease expectations would misrepresent delivered functionality and divert scarce energy budget. Instead, the project documents alternative tracks (Option B QIM watermarking, Option C PyO3/Rust port) in ADRs, clearly labeling them as future work contingent on hardware funding and bench instrumentation.

Equipment discussions are treated transparently: new hardware (oscilloscope, FPGA stimulus, bio-impedance rigs) must map to a concrete experiment plan tied to documented milestones. Until funding is committed, Track1 prioritizes the achievable Option A pipeline and avoids promising features (``watermark'' keywords) that are not implemented. This preserves integrity of reporting and maintains trust with community partners who rely on verifiable telemetry rather than speculative features.
