\subsection{Internet-Draft Submission}
\label{subsec:internet-draft}

As a direct output of this project, we prepared an \ac{IETF} Internet-Draft
(\texttt{draft-trackone-verifiable-telemetry-ledgers-00}) specifying the
verifiable telemetry ledger profile described in this report.
The draft
targets the Independent Submission stream as an Informational document
and was validated with \texttt{xml2rfc}~v3 and \texttt{idnits}.

\subsubsection{Scope and Positioning}

The draft codifies the pipeline semantics developed in this project ---
canonical fact encoding, deterministic Merkle commitment, day chaining,
and OpenTimestamps anchoring --- into a self-contained specification
suitable for independent implementation.
It explicitly positions itself
as complementary to active \ac{IETF} work:

\begin{itemize}
    \item \textbf{SCITT} (draft-ietf-scitt-architecture): generic
    supply-chain transparency; our profile addresses the
    disconnected, daily-batching case S\ac{CI}TT does not cover.
    \item \textbf{COSE Merkle Tree Proofs} (draft-ietf-cose-merkle-tree-proofs):
    \ac{CBOR}-encoded proof primitives; our implementation includes a
    \ac{CBOR} migration path via \texttt{trackone-core}.
    \item \textbf{RATS}: device attestation; deferred in our profile to
    the provisioning layer.
\end{itemize}

\subsubsection{Implementation Status}

The draft references the published Rust crates (\texttt{trackone-core},
\texttt{trackone-ledger}, and others on \texttt{crates.io}) and the
Python gateway as a reference implementation per \ac{RFC}~7942.
The pipeline
has been validated end-to-end on an STM32L476RG development board with
emulated telemetry links; no production field deployment has been
conducted at the time of writing.

\subsubsection{Disclosure}

Early structural drafts of the Internet-Draft were prepared with AI
writing assistance.
All technical content and design decisions reflect
the author's work as documented throughout this report and the project's
architectural decision records.
This is disclosed in the draft's
Acknowledgments section, consistent with \ac{IETF} guidance on AI-assisted
contributions.
