\section{Introduction}

Heritage site monitoring presents unique constraints: sensors must operate for years without maintenance, avoid invasive installation, and preserve data integrity across decades without relying on commercial vendors that may discontinue services.
While modern IoT platforms provide telemetry infrastructure, they typically require vendor lock-in (AWS IoT, Azure IoT) or incur per-message costs (blockchain-native systems like Helium Network).

Track1 addresses a specific niche: \textbf{publicly-auditable telemetry for resource-constrained heritage sensors} where multi-decade autonomy is required and vendor lock-in is unacceptable.
The system combines ultra-low power LoRa communication (<1kB/day traffic budget), edge AEAD encryption (gateway verifies without cloud dependency), and zero-marginal-cost public timestamping via Bitcoin blockchain through OpenTimestamps.

\subsection{Problem Statement}

Heritage site conservators require telemetry systems that:
\begin{itemize}[nosep]
    \item Operate 2+ years on coin-cell batteries (1.4 Wh/day budget)
    \item Provide cryptographic data integrity (confidentiality + authenticity)
    \item Enable independent verification without trusting gateway operator
    \item Survive vendor discontinuation (no proprietary cloud dependencies)
    \item Minimize operational costs (zero per-message fees)
\end{itemize}

Existing solutions fail one or more requirements:
\begin{itemize}[nosep]
    \item \textbf{Commercial IoT platforms} (AWS IoT Core, Azure IoT Hub): Vendor lock-in, audit logs controlled by provider, pay-per-message pricing
    \item \textbf{Blockchain-native systems} (Helium Network): Per-packet fees (\$0.00001/packet = \$3.65/year for 1 msg/hour), no power constraints
    \item \textbf{Proprietary timestamping} (Guardtime KSI): Enterprise licensing, closed-source verification, not Bitcoin-anchored
    \item \textbf{Certificate Transparency} (RFC 6962): Designed for TLS certificates, not IoT telemetry workflows
\end{itemize}

\subsection{Contributions}

This work presents Track1, an \textbf{open-source implementation} combining low-power cryptography, deterministic Merkle batching, and Bitcoin timestamping for heritage site telemetry.
While individual components are well-established (AEAD encryption, Merkle trees, OpenTimestamps), Track1 is the \textbf{first open system} integrating them for ultra-low power LoRa deployments with public Bitcoin verification.

\textbf{Key contributions:}

\begin{enumerate}[nosep]
    \item \textbf{AEAD-protected LoRa frames} (XChaCha20-Poly1305)
    \begin{itemize}[nosep]
        \item Confidentiality + authenticity in <1kB/day traffic budget
        \item Replay protection via sliding window (64 frames, persistent state)
        \item Compatible with LoRaWAN duty cycle limits (1\% EU868)
        \item Gateway-based verification (no cloud required)
    \end{itemize}

    \item \textbf{Daily Merkle anchoring} via OpenTimestamps
    \begin{itemize}[nosep]
        \item Amortizes blockchain costs: 1 anchor/day vs.
        1000+ individual timestamps
        \item Public verifiability without vendor lock-in (Bitcoin blockchain)
        \item Reproducible roots: canonical JSON (RFC 8785 style), sorted leaves
        \item Independent verification CLI for third-party auditors
    \end{itemize}

    \item \textbf{Heritage deployment constraints}
    \begin{itemize}[nosep]
        \item 2+ year battery life on CR2032 (220 mAh @ 3V)
        \item Non-invasive installation (adhesive mount, no wiring)
        \item Open protocols (no proprietary dependencies)
        \item Zero marginal cost per telemetry record
    \end{itemize}

    \item \textbf{Production-ready implementation}
    \begin{itemize}[nosep]
        \item PyNaCl-based gateway with 86\% test coverage (181 tests)
        \item Comprehensive ADRs documenting design decisions
        \item CI-verified OTS workflow (Bitcoin Core headers-only mode)
        \item Real Bitcoin attestation: Block 919384 (October 16, 2025)
    \end{itemize}
\end{enumerate}

\subsection{Comparison with Existing Systems}

\begin{table}[h]
\centering
\caption{Track1 vs. existing IoT and timestamping systems}
\begin{tabular}{lcccc}
\hline
\textbf{System} & \textbf{Public Audit} & \textbf{Cost/Day} & \textbf{Power Budget} & \textbf{Open Source} \\
\hline
AWS IoT Core & Vendor logs & \$0.50 & None & No \\
Azure IoT Hub & Vendor logs & \$0.25 & None & No \\
Helium Network & Blockchain & \$0.09 & None & Partial \\
Guardtime KSI & Proprietary & Enterprise & None & No \\
Cert.\ Transparency & Public (CT logs) & N/A & N/A & Yes \\
\textbf{Track1} & \textbf{Bitcoin (OTS)} & \textbf{\$0.001} & \textbf{1.4 Wh} & \textbf{Yes (MIT)} \\
\hline
\end{tabular}
\end{table}

\textbf{Novelty claim:} While Certificate Transparency (RFC 6962) and Guardtime KSI demonstrate Merkle timestamping at scale, and blockchain systems like Helium provide decentralized IoT infrastructure, Track1 is the \textbf{first open-source implementation} combining:
\begin{itemize}[nosep]
    \item Ultra-low power radio (LoRa, <1kB/day)
    \item Edge AEAD encryption (gateway verifies, no cloud)
    \item Zero-cost public auditability (Bitcoin via OTS, not per-message fees)
\end{itemize}

This targets a \textbf{specific niche}: heritage sites requiring publicly-auditable telemetry where vendor lock-in is unacceptable and multi-decade sensor autonomy is mandatory.

\subsection{Real-World Impact}

Heritage site conservators think in \textbf{decades}, not product cycles.
Track1 addresses long-term concerns:

\begin{itemize}[nosep]
    \item \textbf{Vendor discontinuation:} AWS can deprecate IoT services (e.g., AWS IoT 1-Click discontinued 2020).
    Bitcoin blockchain persists 100+ years.
    \item \textbf{Pricing changes:} Cloud providers can increase fees.
    OTS anchoring costs are fixed (Bitcoin transaction fees amortized across all users).
    \item \textbf{Data sovereignty:} Museums require independent verification.
    Track1 verification CLI runs offline with pre-downloaded Bitcoin headers.
    \item \textbf{Open protocols:} Complete system reproducible from source code.
    No proprietary black boxes.
\end{itemize}

\subsection{Document Structure}

Section 2 covers threat model and requirements.
Section 3 describes system architecture.
Section 4 details cryptographic design (XChaCha20-Poly1305, HKDF, nonce construction).
Section 5 presents pipeline implementation (frame verification, Merkle batching, OTS anchoring).
Section 6 documents schemas and canonicalization policy.
Section 7 validates the system through experiments.
Section 8 presents results (192 tests, 84\% coverage, Bitcoin block 919384 attestation).
Section 9 discusses related work and ethical considerations.
Section 10 concludes with roadmap.

\textbf{Current Status (v0.0.1-m4, October 22, 2025):}
Production XChaCha20-Poly1305 AEAD, reproducible Merkle roots, real Bitcoin OTS anchoring verified, and comprehensive test suite (181 tests passing, 84\% coverage across gateway and pod\_sim modules).
All ADRs accepted through ADR-009.

\subsection{Competing Directions and Deferred Work}
\textbf{Why not image watermarking or steganography?} Track1 handles scalar telemetry (temperature, RH, strain) emitted by pods without onboard cameras. Embedding image watermarks would burn energy budget on sensors that have no imaging pipeline, add transmission overhead, and fail to address the actual threat model (forged telemetry records). Therefore, steganography remains out-of-scope until image-bearing payloads exist.

\textbf{Why not QIM on bio-impedance now?} Option B (Quantization Index Modulation watermarking on bio-impedance) requires lab instrumentation, deterministic stimulus, and higher bandwidth links. It is documented as future research (see ADR~\ref{adr:qim-watermarking}) but deliberately deferred until Track1's telemetry ledger reaches v0.1.0 maturity and dedicated hardware funding is secured.

\textbf{Why not Rust/PyO3 migration yet?} Option C (PyO3 bridge to a Rust core, ADR-017) promises tighter power envelopes and better determinism, but increases build complexity and maintenance overhead. The current milestone keeps the Python gateway for agility; Rust modules remain a latent roadmap goal tied to post-0.1.0 benchmarking (ADR~\ref{adr:benchmarking}) and hardware refresh.

In summary, Track1 prioritizes \textbf{Option A}: verifiable telemetry for heritage sensors under severe power and connectivity constraints. Alternative tracks are catalogued as future work with explicit ADRs and will be revisited once the foundational ledger pipeline is fully productized.
