\section{System Requirements}
\label{sec:system-requirements}

This section captures the functional requirements, non-functional requirements, and constraints for the Barnacle Pod gateway and ledger system.
Each requirement is written in a verifiable form and avoids prescribing implementation details unless mandated by external constraints.

\subsection{Functional Requirements}
\label{subsec:functional-requirements}

\begin{description}[style=nextline]
  \item[FR-1 Ingest sensor frames] The system shall ingest radio frames from all configured sensors and associate each frame with a unique device identifier.

  \item[FR-2 Anti-replay window] The system shall maintain, per device, a replay window over a frame counter, tracking recently seen counter-values to enforce replay policy.
  Frames within a configured bidirectional sliding window that have not been seen before shall be accepted (including modest out-of-order delivery); duplicate or out-of-window frames shall be rejected.

  \item[FR-3 Ledger writing] The system shall write accepted facts to a daily ledger file (\texttt{day.bin}) in a format suitable for Merkle tree construction and external verification.

  \item[FR-4 Audit logging] The system shall emit audit logs for all received frames, including rejected, malformed, or replayed frames, to a separate audit channel (e.g., syslog or metrics).

  \item[FR-5 Merkle batch generation] The system shall periodically batch ledger entries into a Merkle tree, producing a root hash and associated metadata for each day.

  \item[FR-6 OTS stamping] The system shall produce an external timestamping (\ac{OTS}) proof for each daily Merkle root and store that proof alongside the corresponding ledger artifacts.

  \item[FR-7 Verification tooling] The system shall provide a command-line tool that verifies a given \texttt{day.bin} and its associated \ac{OTS} proof, failing verification if any ledger entry has been modified or removed.

  \item[FR-8 Configuration management] The system shall load device configuration (keys, salts, IDs) from a well-defined configuration file and shall fail safely if the configuration is missing or inconsistent.

  \item[FR-9 Error handling] The system shall detect and report configuration, decryption, and I/O errors and shall not silently drop frames or modify ledger data without emitting an error.

  \item[FR-10 Stationary calendar anchoring] The system SHALL support a stationary \ac{OTS} calendar for \ac{CI} and long-running deployments.
  Every \texttt{day.bin} produced by the pipeline SHALL either: (a) get an \ac{OTS} proof within a bounded delay (e.g., 24~h), or (b) be explicitly marked as ``pending'' and retried until either success or operator intervention.

  \item[FR-11 Anti-replay invariants] For each \texttt{(device\_id, frame\_counter)} pair, at most one Fact \ac{SHA}LL be accepted into the Merkle set for a given day.
  Duplicate frames or out-of-window counters MUST be rejected by the Verifier and visible in audit logs but NOT present in \texttt{day.bin} or the day Merkle tree.

  \item[FR-12 Cross-implementation equivalence] The Rust TrackOne-core implementation \ac{SHA}LL produce bit-for-bit identical results to the Python reference for: Fact serialization, Merkle root computation, and \ac{AEAD} encryption/decryption for a canonical golden corpus of test inputs.

  \item[FR-13 Multi-ISA determinism] TrackOne-core SHALL produce identical Facts, Merkle roots, and \ac{AEAD} ciphertexts
  across supported \ac{ISA} targets (e.g., ARM Cortex-M0+/M4, RISC-V softcores) for a shared golden corpus.

  \item[FR-14 Deterministic commitment bytes] For each Fact written into \texttt{day.bin}, the system \ac{SHA}LL define a
  deterministic commitment-byte encoding (TrackOne Canonical \ac{CBOR}) and \ac{SHA}LL test that the commitment bytes are
  stable across code changes and platforms.
  Commitment bytes MUST be treated as the input to Merkle hashing.

  \item[FR-15 Deterministic time encoding] Facts SHALL store time fields as integer epoch seconds (\ac{UTC}).
  Any export to external \ac{API}s (e.g., SensorThings) SHALL deterministically render these times as \ac{RFC}3339 \ac{UTC} strings.
\end{description}

\subsection{Non-Functional Requirements}
\label{subsec:non-functional-requirements}

\begin{description}[style=nextline]
  \item[NFR-1 Integrity] The system shall guarantee that any modification of \texttt{day.bin} or its associated metadata after \ac{OTS} stamping is detectable by the verification tooling.

  \item[NFR-2 Separation of concerns] The system shall maintain a clean separation between the ledger (truth) and audit logs (noise, attacks, malformed input).
  No rejected or replayed frames shall appear in the ledger.

  \item[NFR-3 Performance] The system shall be capable of processing at least $N$ frames per second (to be specified per deployment) on the target hardware while maintaining anti-replay and ledger invariants.

  \item[NFR-4 Dependability] Under normal operating conditions (no disk failure, no configuration corruption), the system shall not lose accepted frames; any I/O failure shall be surfaced as an explicit error.

  \item[NFR-5 Testability] Each requirement in this section shall be verifiable by at least one of: automated test, inspection, or analysis.
  Vague requirements (e.g., ``fast'') shall be refined into measurable criteria.

  \item[NFR-6 Maintainability] Architectural decisions (e.g., anti-replay policy, ledger vs.\ audit separation) shall be documented as Architecture Decision Records (\ac{ADR}s) and kept consistent with the implementation and tests.

  \item[NFR-7 Anchoring resilience] The system SHOULD remain verifiable when public calendars are unavailable, by relying on a stationary calendar and later reconciliation proofs.

  \item[NFR-8 Implementation consistency] Divergence between Python and Rust implementations MUST be detectable via automated tests in \ac{CI}, and \ac{SHA}LL fail the pipeline.
\end{description}

\subsection{Constraints}
\label{subsec:constraints}

\begin{description}[style=nextline]
  \item[C-1 Platform] The gateway software shall run on a Linux-based system using Python as the primary implementation language (with Rust extensions for core logic).

  \item[C-2 Cryptography] The system shall use approved cryptographic primitives and libraries as specified by project policy (XChaCha20-Poly1305, Ed25519, \ac{SHA}-256), and shall not implement bespoke cryptographic algorithms.

  \item[C-3 Counter wrap-around] For the Barnacle Pod deployment profile, 32-bit frame counter wrap-around is considered out-of-policy due to its negligible probability within the system lifetime.
  Any observed wrap-around shall be treated as a provisioning or firmware error.

  \item[C-4 Storage format] Daily ledger files (\texttt{day.bin} and associated \ac{JSON}/metadata) shall be written in backward-compatible formats once deployed, or accompanied by a documented migration path.

  \item[C-5 Logging and privacy] Audit logs shall not contain sensitive key material.
  Device identifiers and counters may be logged; secret keys and seeds shall never be logged in plaintext.
\end{description}

\subsection{Project Scope and Roadmap}\label{subsec:project-scope-and-roadmap}
TrackOne is explicitly framed as ``Track One of Three'': the core pipeline that provides ingestion, cryptographic integrity, deterministic batching, and public anchoring.
It is an intentionally narrow, verifiability-first track whose deliverables are:
\begin{itemize}[nosep]
  \item A hardened ingestion pipeline (framed ND\ac{JSON} + \ac{AEAD} semantics) and replay protection;
  \item Deterministic canonicalization and Merkle batching with day-chunked artifacts;
  \item Public anchoring (OpenTimestamps) and verifier tooling that lets third parties recompute and attest daily roots;
  \item Operational guidance and minimal hardware reference that satisfy the digital contract (frame shape, nonce construction, provisioning).
\end{itemize}

Features reserved for later tracks (Track Two / Three) include field-deployable control loops, sophisticated on-pod actuation safety interlocks, large-scale analytics and UI, and long-running automated key-rotation/\ac{PKI} operations.
TrackOne is positioned as an \ac{MVP} that is ``production-informed'' but not yet a full safety-certified control system.

\subsection{Heritage deployment tracks}\label{subsec:heritage-deployment-tracks}
TrackOne is explicitly positioned as \textbf{Track One of Three}.
Each track increases scope while preserving the verifiability guarantees:
\begin{itemize}[nosep]
  \item \textbf{Track One (this report):} Evidence pipeline for sensor pods---\ac{AEAD} framing, replay protection, canonical facts, Merkle batching, Bitcoin anchoring (\ac{OTS}), and independent verification tooling.
  Hardware requirements stop at the digital contract (nonce construction, sealed facts).
  \item \textbf{Track Two (SensorThings, LoRa, and Operational Intelligence):} Extension to \ac{OGC} SensorThings \ac{API}s (\ac{ADR}s 027--030), adaptive \ac{LoRa} uplink policies (\ac{ADR}-025), and \ac{GIS}-informed placement (\ac{ADR}-031).
  This track introduces the stationary calendar service (\ac{ADR}-022) and the virtual fleet validation framework (\ac{ADR}-033).
  \item \textbf{Track Three (Rust core, PQC, and Actuation Safety):} Rust first-class gateway crates (maturin exports) replace Python hot paths (\ac{ADR}-017, \ac{ADR}-019).
  Post-quantum hybrid provisioning (\ac{ADR}-036), OTA firmware safety (\ac{ADR}-026), and signature roles (\ac{ADR}-037).
  Embedded Rust firmware (STM32L0/L4, RISC-V) with adaptive sampling.
\end{itemize}
This scoping keeps requirements concrete and prevents speculative features from diluting the verifiability focus of Track One.

% Reviewer-slim version: omit domain-specific operational constraints.
\subsection{Operational and Environmental Constraints}\label{subsec:operational-and-environmental-constraints}
For subterranean irrigation deployments, the physical environment and service model materially change requirements.
We therefore set explicit operational constraints that must be observed when deploying TrackOne artifacts in the field:
\begin{itemize}[nosep]
  \item \textbf{Deployment tolerance:} Pods are expected to be wet, buried, and inaccessible for long durations; hardware must therefore favour robustness and conservative scheduling for maintenance windows.
  \item \textbf{Maximum allowed downtime:} Configurable per site (typical value 24 hours); longer outages must be escalated to human operators.
  \item \textbf{Acceptable data loss:} Define a site-specific bound (e.g., no more than 0.5\% of day's facts lost) beyond which operator rollback or manual audit is required.
  \item \textbf{Sampling frequency:} Default 1--10 samples per hour per pod depending on sensor; higher rates must be justified by an explicit risk/benefit analysis.
  \item \textbf{Firmware update policy:} Updates must support atomic rollback and require an operator-confirmed maintenance window; over-the-air auto-update that can change actuation logic is out-of-scope for TrackOne.
  \item \textbf{Storage limits:} On-device and gateway storage quotas must be set; when storage is full, pods/gateway must stop non-essential writes and raise an alert to operator channels rather than silently dropping telemetry.
  \item \textbf{Stationary calendar usage:} Use of the stationary calendar in \ac{CI} and weekly ``ratchet'' jobs is MANDATORY when public \ac{OTS} pools are unstable.
  \item \textbf{Embedded viability:} At least one Cortex-M0+/M4 and one RISC-V firmware build \ac{SHA}LL be kept
  in-tree and exercised regularly (stress harness or FPGA target) to guarantee that TrackOne-core behaves identically
  across \ac{MCU} families.
  \item \textbf{GIS-constrained placement:} Sensor pod placements \ac{SHA}LL be drawn from pre-approved \ac{GIS} polygons (see Section~\ref{sec:site_planning}), not arbitrary coordinates.
\end{itemize}

\subsection{GIS-Informed Pod Placement Constraints}\label{subsec:gis-informed-pod-placement-constraints}
For subterranean irrigation corridors (khettara-like tunnels) and adjacent heritage structures, pod placement must be planned with basic geospatial constraints that can be encoded in deployment runbooks:
In water-domain deployments, the site plan and placement constraints also support governance and reporting obligations (Section~\ref{subsec:water-law}).
\begin{itemize}[nosep]
  \item \textbf{Buffer zones:} Maintain minimum stand-off distances from protected masonry, inscriptions, and water intakes (site-specific, e.g., 30--50~cm) to preserve reversibility and avoid damage.
  \item \textbf{Radio propagation:} Prefer line-of-sight along shafts; where bends or thick walls exist, introduce relay pods at bends mapped in \ac{GIS} to keep link budgets within margins.
  \item \textbf{Hydrology overlays:} Avoid low points identified by flood maps for primary pods; place flood sentinels explicitly at risk points to trigger maintenance/evacuation procedures.
  \item \textbf{Access waypoints:} Every pod should be within a serviceable reach from registered maintenance paths; the \ac{GIS} plan must list approach routes and safe access times (seasonal constraints).
  \item \textbf{Cultural heritage constraints:} Mounts must be non-invasive (adhesive/straps) and removable; no drilling or permanent fixtures without curator approval.
\end{itemize}

\subsection{Spatial Database Requirements}\label{subsec:spatial-database-requirements}
To keep \ac{GIS} reasoning close to the verifiable pipeline while avoiding heavyweight external services, TrackOne MAY use a
Spatialite-backed SQLite database for pod placement, maintenance routes, and hydrology overlays.
We require:

\begin{description}[style=nextline]
  \item[SD-1 Single source of spatial truth] Site polygons, buffer zones, and approved maintenance routes \ac{SHA}LL be
  stored in a version-controlled Spatialite database.
  Generated pod placement plans MUST reference this database
  rather than ad-hoc shapefiles.
  \item[SD-2 Read-side separation] The ingest and ledger paths \ac{SHA}LL treat the Spatialite database as a read-only state.
  Failure or corruption of spatial indices MUST NOT compromise ledger integrity; it can only degrade planning and
  visualization.
  \item[SD-3 Export compatibility] Spatialite contents \ac{SHA}LL be exportable to standard \ac{GIS} formats (e.g., GeoPackage,
  Shapefile) for use in Q\ac{GIS}/Arc\ac{GIS} without loss of topology or metadata relevant to pod placement.
\end{description}

\subsection{Threat Model}
\label{subsec:threat-model}

We assume TrackOne is deployed in adversarial and failure-prone environments (remote sites, intermittent access, mixed trust on backhaul).
The key objective is that the canonical ledger artifacts (\texttt{day.bin} + Merkle metadata + \ac{OTS} proofs) remain \emph{verifiable} and \emph{non-equivocable} even if upstream networks, ingest inputs, or operational tooling degrade.

We categorize threats and failure modes along the following axes.
Each threat is paired with explicit TrackOne countermeasures and verification hooks.

\begin{description}[style=nextline]
  \item[T-1 Replay and ledger pollution] Attackers or misbehaving devices may replay, duplicate, or reorder frames to confuse operators and attempt to pollute the ledger.
  \textbf{Countered by:} FR-2, FR-3, FR-11, NFR-2; ledger vs.
  audit separation; and tests in \texttt{test\_replay\_*.py}.

  \item[T-2 Timestamp and calendar manipulation] \ac{OTS} calendars (public or stationary) may be unavailable, inconsistent, or malicious, leading to incomplete, delayed, or misleading proofs.
  \textbf{Countered by:} FR-6, FR-10, NFR-1, NFR-7, RR-1--RR-3; stationary calendar fallbacks; and the weekly ratchet job that upgrades pending proofs.

  \item[T-3 Commitment-encoding drift (transport vs. commitment bytes)] Ambiguity about which bytes are hashed (wire/transport encoding vs.
  canonical commitment encoding) can create silent integrity splits: roots that verify in one toolchain but cannot be reproduced by auditors.
  \textbf{Countered by:} \ac{ADR}-034; a deterministic TrackOne Canonical \ac{CBOR} profile for commitment bytes (distinct from Postcard transport bytes); and stability tests in \texttt{crates/trackone-core/src/cbor.rs}.

  \item[T-4 Cross-implementation divergence (Rust/Python/ISA)] Different gateway implementations (Rust vs.
  Python), or different CPU targets (Cortex-M vs.
  RISC-V), may disagree on Fact encoding, Merkle roots, or \ac{AEAD} semantics.
  \textbf{Countered by:} FR-12, FR-13, NFR-8; golden-corpus tests; and TrackOne-core determinism tests across targets.

  \item[T-5 Key compromise and operator-plane misuse] Theft of configuration, weak operational practices, or unauthorized operator actions could allow:
  (a) forging frames, (b) impersonating devices, or (c) suppressing/rewriting non-anchored artifacts.
  \textbf{Countered by:} C-5 (no secret logging), FR-8/FR-9 (fail-safe config), separation of duties (RR-4), immutable day artifacts once written, and explicit operator action logging (R-5).

  \item[T-6 Backhaul interception and traffic analysis] Gateways may use untrusted uplinks (public Wi-Fi/cell), enabling interception, tampering, and traffic analysis between gateway and master server.
  \textbf{Countered by:} authenticated transport hardening (e.g., a to-be-deployed WireGuard backhaul) as an operational control; and TrackOne's end-to-end artifact integrity (Merkle + \ac{OTS}) which remains verifiable even if transport confidentiality fails.
  WireGuard mitigates backhaul interception/tampering but does not protect against RF-layer attacks that occur before the gateway.

  \item[T-7 Time semantic confusion (phenomenon vs. ingest/result time)] If time fields are ambiguous, mixed, or emitted in inconsistent formats/timezones, downstream projections (SensorThings/Arc\ac{GIS}) can misrepresent events, and auditors can misinterpret the measurement window.
  \textbf{Countered by:} SA-2; explicit time-field mapping rules; and the requirement that all operator-facing timestamps be expressed as \ac{RFC}3339 (\ac{UTC}) while retaining the raw integer epoch seconds in Facts.

  \item[T-8 Identifier ambiguity (PodId formatting)] If device identifiers have inconsistent textual renderings (endianness, truncation, or mixed hex conventions), operators may mis-attribute telemetry and incident response may target the wrong pod.
  \textbf{Countered by:} fixed-width \texttt{PodId} (8 bytes) in the Fact schema; stable \texttt{PodId} conversion from legacy 32-bit IDs; and consistent hex rendering rules in tooling.

  \item[T-9 Crypto agility] ``Harvest-now, decrypt-later'' is a risk for any confidentiality-protected
  provisioning messages captured today.
  \textbf{Countered by:} suite negotiation and transcript binding in provisioning designs (hybrid-ready), while allowing pods to remain X25519-only in early deployments. \textbf{Downgrade resistance:} any negotiated \texttt{kex\_suite} MUST be authenticated and bound to the provisioning transcript.
  This threatens confidentiality, not ledger integrity; telemetry is dominated by radio airtime, not by suite choice.

  \item[T-10 Denial-of-service and resource exhaustion] Attackers may attempt to reduce availability or consume duty-cycle budget via RF jamming, replay floods, bogus-device floods, malformed frame floods, or deliberate triggering of rejection paths (log noise).
  \textbf{Countered by:} FR-4 (audit logging), NFR-2 (ledger vs.\ audit separation), and operational controls (rate limiting, allowlists, monitoring).
  DoS affects timeliness and availability; it does not invalidate the integrity of anchored artifacts.
\end{description}

\textbf{Key lifecycle note (rotation / revocation / quarantine):}
Track One treats key lifecycle as an operational control rather than a fully automated \ac{PKI} system.
When keys are rotated, the gateway updates the device configuration and resets or reinitializes the per-device session state (e.g., replay window state) as part of the provisioning procedure.
When a pod is suspected compromised, operators may revoke/quarantine it by removing its keys from the active configuration and refusing future frames from that device identifier; such frames may still be retained as audit noise but MUST NOT be admitted into \texttt{day.bin} or the Merkle set.
Automated fleet-wide rotation and long-running revocation workflows are deferred to later tracks, but the boundary is explicit here to avoid implying that cryptography alone provides fleet safety.

A more detailed attacker model and mapping to countermeasures is given in Section~\ref{sec:security}.

% Reviewer-slim version: omit irrigation/SensorThings requirement details.
\subsection{Irrigation-Specific Requirements}\label{subsec:irrigation-specific-requirements}
To support \textbf{traceability} and operational safety in irrigation contexts, we add a small set of testable, measurable requirements that must be in scope for any deployment using TrackOne artifacts:
For Moroccan water-domain deployments, these requirements must be considered alongside regulatory obligations under Loi36--15(Section~\ref{subsec:water-law}).

\begin{description}[style=nextline]
  \item[R-1 Maximum anchoring delay] For any given day, anchors (\ac{OTS} proofs) must be present within 72 hours of day end or the system must mark the day as ``pending anchor'' and notify operators.
  \item[R-2 Maximum tolerable data loss] Daily fact loss must not exceed site-defined limits (default 0.5\%). Exceeding this triggers operator review.
  \item[R-3 Sampling \& action separation] Control decisions that affect irrigation volumes MUST not be taken on facts that are older than $T$ hours (default: 24 h) without human confirmation.
  \item[R-4 Update rollback] Any firmware update that modifies actuation logic must support a verified rollback mechanism and be staged to a small pilot group for at least 7 days before full rollout.
  \item[R-5 Auditability] All critical operator actions (e.g., manual valve overrides, firmware promote) must be logged with operator identity, timestamp, and Merkle-anchored proof of the day's root that covers the action metadata.
\end{description}

\subsection{SensorThings and API Projection Requirements}\label{subsec:sensorthings-and-api-projection-requirements}
TrackOne treats \ac{OGC} SensorThings-style \ac{API}s and MQTT bindings as \emph{projections} of the canonical Fact stream, not
as authoritative storage.
To keep the attack surface narrow while remaining standards-aligned, we require:

\begin{description}[style=nextline]
  \item[SA-1 Canonical-to-API mapping] All SensorThings ``Things'', ``Locations'', ``Datastreams'', and ``Observations''
  exposed by the gateway \ac{SHA}LL be derived from immutable Facts (e.g., \texttt{EnvFact}) that are part of a Merkle
  day-set.
  No \ac{API}-only records may exist without a corresponding Fact.

  \item[SA-2 Phenomenon vs.\ result time] The gateway \ac{SHA}LL map \texttt{EnvFact.phenomenon\_time\_start/end} to
    \texttt{phenomenonTime} and \texttt{Fact.ingest\_time} to \texttt{resultTime} in SensorThings payloads.
  Operator-facing \ac{API} timestamps (including SensorThings \ac{JSON}) \ac{SHA}LL be emitted as \ac{RFC}3339 strings in \ac{UTC}
  (derived from the canonical epoch-second fields), to avoid timezone drift and ``local time'' ambiguity.

  \item[SA-3 Stable type identity] The gateway \ac{SHA}LL map \texttt{EnvFact.sample\_type} to a stable ObservedProperty identity.
    Any change to the semantic meaning of a \texttt{SampleType} (or its unit) MUST trigger a projection version bump.

  \item[SA-4 API hardening] SensorThings REST and MQTT endpoints \ac{SHA}LL validate inputs against strict schemas, enforce
  rate limits, and avoid exposing raw keys or internal identifiers. \ac{API} misuse \ac{SHA}LL not be able to inject or modify
  ledger Facts; at most it can affect cached projections.

  \item[SA-5 Read-only by default] Public SensorThings endpoints \ac{SHA}LL be read-only with respect to the core ledger.
  Any write-capable \ac{API} (e.g., for site metadata) MUST be authenticated and kept off the critical ingest path.
\end{description}

\subsection{Calendar, Anti-Replay, and Implementation Requirements}
\label{subsec:calendar-anti-replay-impl-reqs}
This subsection groups rationale and test hooks for the anchoring, anti-replay, and determinism requirements (FR-10--FR-15, NFR-7--NFR-8) listed in Sections~\ref{subsec:functional-requirements} and~\ref{subsec:non-functional-requirements}.

\subsection{Resilience Requirements}\label{subsec:resilience-requirements}
We add concise, testable targets that express how the pipeline maintains useful service across outages and partial failures:
\begin{description}[style=nextline]
  \item[RR-1 Anchor backlog clearance] After an uplink outage of up to 7 days, gateways shall clear the \ac{OTS} anchoring backlog within 48~hours of connectivity restoration while preserving per-day artifact immutability.
  \item[RR-2 Multi-gateway replication] In multi-gateway deployments, day artifacts shall be replicated to at least one peer within 24~hours (\ac{ADR}-019 Supernode Train) to bound single-node loss.
  \item[RR-3 Stationary calendar fallbacks] When public calendars are unavailable, stationary calendar proofs shall be accepted with a visible \emph{pending public anchor} status in meta until upgraded.
  \item[RR-4 Peer attestation quorum] Day roots shall collect at least $q$ operator signatures (default $q{=}2$ of $n$) before being marked \emph{attested} in operator dashboards.
  \item[RR-5 GIS-linked maintenance] Pods flagged by flood sentinel events or repeated RF failures must enter a \ac{GIS} task list within 24~hours, including route and safety notes for field teams.
\end{description}

\subsection{Energy Budget and Coin-Cell Viability}\label{subsec:energy-budget-and-coin-cell-viability}
For buried pods powered by small primary cells (e.g., CR2032) or harvesters plus a storage element, we explicitly
separate average energy use (mAh / Wh) from peak current viability (ability to sustain \ac{LoRa} TX bursts):
\begin{itemize}[nosep]
  \item \textbf{Assumed cell:} CR2032-class, $\approx 200$~mAh usable at low drain, $E \approx 0.6$~Wh at 3.0~V.
  \item \textbf{Mission duration:} 6~months ($\approx 4320$~h).
  \item \textbf{Sleep current target:} $I_{\text{sleep}} \le 5\,\mu\text{A}$ (\ac{MCU} + leakage + powered-down sensors).
  \item \textbf{Uplink cadence:} 1--4 uplinks/day, each carrying a compact summary (typically $< 80$~B).
  \item \textbf{LoRa TX envelope:} $I_{\text{tx}} \in [40,120]$~mA, airtime 0.5--2.0~s per uplink depending on \ac{SF}/payload.
\end{itemize}
We use the following duty-cycle model for capacity estimates:
\begin{equation}
  E_{\text{wake}} \approx V \cdot (I_{\text{mcu}} t_{\text{mcu}} + I_{\text{sensor}} t_{\text{sensor}} + I_{\text{tx}} t_{\text{tx}} + I_{\text{rx}} t_{\text{rx}}),\label{eq:equation2}
\end{equation}
with total charge over the mission
\begin{equation}
  Q_{\text{total}} \approx I_{\text{sleep}} T_{\text{mission}} + N_{\text{wake}} \cdot \frac{E_{\text{wake}}}{V}.\label{eq:equation}
\end{equation}
Under representative parameters (sleep 5~$\mu$A, 1--4 uplinks/day, $t_{\text{tx}} \le 2$~s) we obtain $Q_{\text{total}} \approx
30$--65~mAh over 6~months: \emph{feasible on a single CR2032 by energy}.
However, peak current and voltage sag during
\ac{LoRa} TX require a reservoir capacitor (or supercap) and in practice motivate either two cells in parallel or a
harvester + storage element.

We therefore state explicit, testable requirements:
\begin{description}[style=nextline]
  \item[ER-1 Coin-cell budget] A reference pod configuration (\ac{MCU} + SHTC3-only sensing, 1 uplink/day, $t_{\text{tx}} \le 1$~s at
  moderate \ac{SF}) \ac{SHA}LL demonstrate a measured average current $I_{\text{avg}} \le 20\,\mu$A over a 7-day endurance test.
  \item[ER-2 Peak-current mitigation] Pods using CR2032-class cells MUST include a local reservoir capacitor sized to
  keep supply voltage within the \ac{MCU} and radio operating bounds during \ac{LoRa} TX peaks.
  Designs that cannot meet this
  \ac{SHA}LL use a different cell chemistry or an energy-harvesting front-end (e.g., BQ25570 + supercap).
  \item[ER-3 Reporting of assumptions] Deployment runbooks \ac{SHA}LL record assumed uplink cadence, \ac{SF},
  and worst-case airtime so that operators can relate conditions back to the energy model.
\end{description}

\subsection{Traceability and Verification}
\label{subsec:requirements-traceability}

Requirements are verified via specific automated tests in the \ac{CI} pipeline:

\begin{table}[H]
\centering
\small
\caption{Requirements Traceability Matrix (Functional)}
\label{tab:req-traceability-fun}
\begin{tabularx}{\textwidth}{|l|>{\raggedright\arraybackslash}X|l|}
\hline
\textbf{Req ID} & \textbf{Requirement Description} & \textbf{Verification Artifact / Test} \\ \hline
FR-1  & Ingest sensor frames           & \texttt{test\_pod\_sim}, \texttt{test\_framed\_ingest} \\ \hline
FR-2  & Anti-replay window             & \texttt{test\_replay\_window\_unit} \\ \hline
FR-3  & Ledger writing                 & \texttt{test\_merkle\_batcher\_sees\_facts} \\ \hline
FR-4  & Audit logging                  & \texttt{test\_framed\_ingest} (log inspection) \\ \hline
FR-5  & Merkle batching                & \texttt{test\_merkle\_determinism} \\ \hline
FR-6  & \ac{OTS} stamping                   & \texttt{test\_ots\_integration} \\ \hline
FR-7  & Verification tooling           & \texttt{verify\_cli} (End-to-End) \\ \hline
FR-8  & Config management              & \texttt{test\_gateway\_pipeline} \\ \hline
FR-9  & Error handling                 & \texttt{test\_framed\_security} \\ \hline
FR-10 & Stationary anchor              & \texttt{tests/e2e/test\_ots\_calendar\_pipeline} \\ \hline
FR-11 & Replay invariants              & \texttt{tests/integration/test\_replay\_merkle} \\ \hline
FR-12 & Cross-impl equivalence         & \texttt{tests/integration/test\_rust\_core\_equiv} \\ \hline
FR-13 & Multi-\ac{ISA} determinism          & Rust suite (ARM/RISC-V) \\ \hline
FR-14 & Deterministic commitments       & \texttt{crates/trackone-core/src/cbor.rs} \\ \hline
FR-15 & Deterministic time             & \texttt{tests/integration/test\_sensorthings} \\ \hline
\end{tabularx}
\end{table}

\begin{table}[H]
\centering
\small
\caption{Requirements Traceability Matrix (Non-Functional and Constraints)}
\label{tab:req-traceability-non-fun}
\begin{tabularx}{\textwidth}{|l|>{\raggedright\arraybackslash}X|l|}
\hline
\textbf{Req ID} & \textbf{Requirement Description} & \textbf{Verification Artifact / Test} \\ \hline
NFR-1 & Integrity                      & \texttt{verify\_cli} (End-to-End) \\ \hline
NFR-2 & Concerns split                 & \texttt{test\_merkle\_batcher\_sees\_duplicate\_free} \\ \hline
NFR-3 & Performance                    & Benchmarks (Section 7.9) \\ \hline
NFR-4 & Dependability                  & \ac{CI} Stress tests \\ \hline
NFR-5 & Testability                    & This matrix and 85\% coverage \\ \hline
NFR-6 & Maintainability                & \ac{ADR} inventory (Section 8.3) \\ \hline
NFR-7 & Anchoring resilience           & \ac{CI} ratchet (stationary calendar) \\ \hline
NFR-8 & Implementation consistency      & Rust + Python golden corpus \\ \hline
C-1   & Platform (Linux/Python)        & \ac{CI} Environment \\ \hline
C-2   & Cryptography (Standard)        & \texttt{test\_crypto\_vectors} \\ \hline
C-5   & Logging (No secrets)           & Bandit scan (\ac{ADR}-009) \\ \hline
ER-1  & Coin-cell budget               & 7-day endurance test report \\ \hline
ER-2  & Peak-current mitigation        & Hardware review / traces \\ \hline
R-1   & Anchoring delay                & \texttt{test\_ots\_retry\_logic} \\ \hline
R-2   & Data loss limit                & Operational monitoring alerts \\ \hline
R-5   & Auditability                   & Merkle-anchored action metadata tests \\ \hline
SA-1/2& SensorThings                   & \texttt{tests/integration/test\_sensorthings} \\ \hline
SD-1/2& Spatialite                     & \texttt{tests/integration/test\_spatialite\_site\_db} \\ \hline
\end{tabularx}
\end{table}
