\section{Results}
\label{sec:results}

This section presents the principal findings from the experiments
described in Section~\ref{sec:experiments}, organised by the claims
they support.
Each subsection states the requirement under test, the
observed outcome, and its implication for the system's readiness.

% ------------------------------------------------------------------
\subsection{Pipeline integrity (FR-1, FR-3, FR-5, NFR-1)}

The end-to-end pipeline — frame emission, \ac{AEAD} verification,
canonical fact writing, Merkle batching, and independent
re-verification — executes without error across all tested
configurations.
Table~\ref{tab:pipeline-summary} summarises the
key invariants validated.

\begin{table}[H]
    \centering
    \caption{Pipeline integrity invariants and observed outcomes.}
    \label{tab:pipeline-summary}
    \small
    \begin{tabularx}{\textwidth}{l X l}
\toprule
\textbf{Invariant} & \textbf{Test method} & \textbf{Result} \\
\midrule
Deterministic Merkle root   & Same facts (shuffled order) $\rightarrow$ identical root
& Pass \\
Byte-identical day blob     & Repeated batching $\rightarrow$ identical \texttt{day.bin} \ac{SHA}-256
& Pass \\
Tamper detection            & Single-bit mutation of \texttt{day.bin} $\rightarrow$ root mismatch
& Pass \\
End-to-end verification     & \texttt{verify\_cli} exit code 0 on valid artifacts
& Pass \\
\bottomrule
    \end{tabularx}
\end{table}

These results confirm that the canonical-fact $\rightarrow$
Merkle-root $\rightarrow$ \ac{OTS}-anchor chain is deterministic and
tamper-evident, satisfying NFR-1 (any post-anchor modification is
detectable).

% ------------------------------------------------------------------
\subsection{Anti-replay enforcement (FR-2, FR-11, NFR-2)}

The sliding replay window (size~64) correctly enforces the
\texttt{(dev\_id,~fc)} single-consumption invariant across all tested
scenarios:

\begin{center}
    \small
    \begin{tabular}{lll}
        \toprule
        \textbf{Scenario} & \textbf{Expected} & \textbf{Observed} \\
        \midrule
        Monotonic increasing \texttt{fc}           & Accept  & Accept  \\
        Duplicate frame (same \texttt{fc})         & Reject  & Reject  \\
        Out-of-window jump ($> 64$)                & Reject  & Reject  \\
        Reordered frames within window             & Accept  & Accept  \\
        Duplicate after gateway restart            & Reject  & Reject  \\
        Boundary: highest $+$ window              & Accept  & Accept  \\
        Boundary: highest $+$ window $+ 1$        & Reject  & Reject  \\
        \bottomrule
    \end{tabular}
\end{center}

Rejected frames appear only in audit logs; they never enter
\texttt{day.bin} or the Merkle set.
This validates the ledger-vs-audit
separation required by NFR-2.

% ------------------------------------------------------------------
\subsection{Cryptographic correctness (C-2)}

All cryptographic primitives pass round-trip and negative tests:

\begin{itemize}[nosep]
  \item \textbf{AEAD:} XChaCha20-Poly1305 encrypt/decrypt matches
  deterministic test vectors; single-bit ciphertext, tag, and
  \ac{AAD} mutations are rejected.
  \item \textbf{Key agreement:} X25519 shared secrets match between
  independently generated key pairs.
  \item \textbf{KDF:} \ac{HKDF}-\ac{SHA}256 outputs match \ac{RFC}~5869 test vectors
  and edge cases (empty salt, zero-length info).
  \item \textbf{Signatures:} Ed25519 sign/verify succeeds for valid
  inputs; tampered messages and wrong keys are rejected.
\end{itemize}

Property-based testing (Hypothesis) found no encoding failures across
$>$10\,000 generated TLV inputs.

% ------------------------------------------------------------------
\subsection{Cross-implementation equivalence (FR-12, NFR-8)}

The Rust \texttt{trackone-core} crate and the Python gateway produce
bit-for-bit identical outputs for a shared golden corpus:

\begin{itemize}[nosep]
  \item Canonical \ac{CBOR} commitment bytes: identical across Python and Rust.
  \item Merkle root computation: identical for all corpus days.
  \item \ac{AEAD} encrypt/decrypt: identical ciphertexts and tags.
\end{itemize}

No divergences were observed. \ac{CI} enforces this invariant: any
cross-implementation mismatch fails the build.

% ------------------------------------------------------------------
\subsection{Embedded viability (FR-13)}

Firmware stress tests confirm that the frame builder fits within
STM32L0-class resource budgets:

\begin{center}
    \small
    \begin{tabular}{llrl}
        \toprule
        \textbf{Target} & \textbf{Iterations} & \textbf{Max frame} & \textbf{Stack} \\
        \midrule
        Host (x86\_64)                & 5.6\,M/s   & 56\,B & $\sim$1\,KiB \\
        QEMU Cortex-M4               & 1\,000      & 56\,B & guard intact \\
        QEMU Cortex-M0+ (thumbv6m)   & 100         & 56\,B & 256\,B guard intact \\
        \bottomrule
    \end{tabular}
\end{center}

Code size ($\sim$3.5\,KiB \texttt{.text}) and RAM usage are within
STM32L072 budgets (128\,KiB flash, 20\,KiB RAM). Wire format is
stable across architectures.
Note: stress binaries use a dummy \ac{AEAD}
to isolate framing behaviour; cryptographic throughput is not claimed.

% ------------------------------------------------------------------
\subsection{OTS anchoring and verification (FR-6, NFR-1)}

Real Bitcoin attestations were obtained and verified locally:

\begin{center}
    \small
    \begin{tabular}{ll}
        \toprule
        \textbf{Property} & \textbf{Value} \\
        \midrule
        Artifact                    & \texttt{day/2025-10-07.bin} \\
        Bitcoin blocks attesting    & 919\,384;\; 919\,388;\; 919\,418 \\
        Calendars used              & \texttt{a.pool.opentimestamps.org} (+ others) \\
        Verification method         & Pruned Bitcoin Core, headers-only ($<$1\,GB) \\
        Merkle root match           & Confirmed \\
        \bottomrule
    \end{tabular}
\end{center}

The day artifact's \ac{SHA}-256 digest matches the sidecar metadata; the
\ac{OTS} proof validates against Bitcoin block headers.
This demonstrates
that the full anchoring lifecycle (stamp $\rightarrow$ upgrade
$\rightarrow$ verify) is functional and that artifacts are
independently verifiable without trusting the gateway operator.

% ------------------------------------------------------------------
\subsection{Test coverage and CI}

\begin{center}
    \small
    \begin{tabular}{lr}
        \toprule
        \textbf{Metric} & \textbf{Value} \\
        \midrule
        Total tests (Python)        & 200+ (181 last run; 2 skipped) \\
        Coverage (gateway + pod)    & $\sim$85\% (target: 85\%) \\
        Python versions tested      & 3.11, 3.12, 3.13 \\
        Static analysis             & mypy, ruff, bandit — all clean \\
        Full suite wall time        & $\sim$20\,s \\
        \bottomrule
    \end{tabular}
\end{center}

Uncovered code is concentrated in \ac{CLI} entry points and rare I/O error
branches.
Optional integration tests (real \ac{OTS}, \ac{TSA} verification)
auto-skip when external tools are unavailable, avoiding false
negatives while keeping real-world checks reproducible.

% ------------------------------------------------------------------
\subsection{Storage and performance}

For a representative day (10 devices, 100 frames each):

\begin{center}
    \small
    \begin{tabular}{lr}
        \toprule
        \textbf{Artifact class} & \textbf{Size} \\
        \midrule
        Frames (ND\ac{JSON})         & $\sim$200\,KB \\
        Facts (canonical)       & $\sim$150\,KB \\
        Block headers + day blob& $\sim$25\,KB \\
        \ac{OTS} proof (confirmed)   & $\sim$10\,KB \\
        \midrule
        \textbf{Total per day}  & \textbf{$\sim$385\,KB} \\
        \bottomrule
    \end{tabular}
\end{center}

Storage scales linearly with device count and sampling rate.
At
the target duty cycle (4 uplinks/day, 10 devices), annual storage
is under 150\,MB — well within the capacity of commodity SD cards
or gateway flash.

% ------------------------------------------------------------------
\subsection{Summary of findings}

Table~\ref{tab:req-verdict} maps the principal requirements to their
experimental outcomes.

\begin{table}[H]
    \centering
    \caption{Requirements traceability: key findings.}
    \label{tab:req-verdict}
    \small
    \begin{tabularx}{\textwidth}{l X l}
\toprule
\textbf{Req.} & \textbf{Claim} & \textbf{Verdict} \\
\midrule
FR-2, FR-11  & Replay window rejects duplicates and stale frames
& Satisfied \\
FR-5, NFR-1  & Deterministic Merkle roots; tamper detected
& Satisfied \\
FR-6         & Daily \ac{OTS} anchor with Bitcoin attestation
& Satisfied \\
FR-7         & Independent verification from published artifacts
& Satisfied \\
FR-12, NFR-8 & Rust/Python cross-implementation equivalence
& Satisfied \\
FR-13        & Firmware fits Cortex-M0+ resource budget
& Satisfied \\
C-2          & Approved primitives; no bespoke crypto
& Satisfied \\
NFR-2        & Ledger/audit separation enforced
& Satisfied \\
\bottomrule
    \end{tabularx}
\end{table}

All Track~One requirements that are testable with the current
(pre-deployment) system are satisfied.
Requirements that depend on
sustained field operation (e.g., R-1 anchoring delay under real
connectivity, R-2 data loss bounds over months) remain to be
validated during a pilot deployment and are identified as future
work.
