\section{Schemas, Merkle Policy, and OTS (ADR-003)}

\subsection{Canonicalization}
Facts and records are serialized as JSON with:
\begin{itemize}[nosep]
  \item UTF-8, sorted keys, no insignificant whitespace.
  \item Stable number/string encoding; no trailing zeros or locale variance.
\end{itemize}
Leaf hash: $h_i=\mathrm{SHA}\text{-}256(\texttt{canonical\_json}(fact\_i))$.

\subsection{Merkle tree}
\begin{itemize}[nosep]
  \item Sort leaf hashes lexicographically to remove input-order dependence.
  \item Pairwise hash up the tree; when odd, duplicate the last hash at that level.
  \item Root is $\texttt{merkle\_root}$; include leaf\_hashes in the header for auditability.
\end{itemize}

\subsection{Block header schema}
Fields: \texttt{version, site\_id, day, batch\_id, merkle\_root, count, leaf\_hashes}. Validated by \texttt{block\_header.schema.json}.

\subsection{Day record schema}
Fields: \texttt{version, site\_id, date, prev\_day\_root, batches[], day\_root}. Canonical day blob written as \texttt{day/YYYY-MM-DD.bin} with a human-readable \texttt{.json} peer. Validated by \texttt{day\_record.schema.json}.

\subsection{OTS anchoring and verification}
\begin{itemize}[nosep]
  \item Anchor the canonical day blob with OpenTimestamps; store proof alongside the blob.
  \item Verification: recompute Merkle root from canonical facts, check against header/day record, then verify OTS proof.
  \item If OTS client unavailable, write a placeholder and defer confirmation; proofs can be upgraded later.
\end{itemize}

\subsection{Schema validation}
When \texttt{--validate-schemas} is enabled, validate against:
\begin{itemize}[nosep]
  \item \texttt{fact.schema.json} (telemetry facts)
  \item \texttt{block\_header.schema.json} (batch header)
  \item \texttt{day\_record.schema.json} (daily aggregates)
\end{itemize}
Validation is warning-only in M\#1 to preserve pipeline progress.
